Species turnover reveals hidden effects of decreasing Nitrogen deposition in mountain hay meadows

### 2018-06-25 Manuscript.Rmd
Corrections of typos.

affiliation:
  - code: 1
    address: Zoological Institute, University of Basel, Basel, Switzerland
  - code: 2
    address: Hintermann  Weber AG, Austrasse 2a, 4153 Reinach, Switzerland
  - code: 3
    address: Meteotest, Fabrikstrasse 14, 3012 Bern, Switzerland
 - code: 4
    address: Agroscope, Swiss Soil Monitoring Network NABO, 8046 Zurich, Switzerland
  - code: 5
    address: Federal Office for the Environment (FOEN), Air Pollution Control and Chemicals Division, 3003 Bern, Switzerland
abstract: >
  Nitrogen (N) deposition is a major threat to biodiversity of many habitats. The recent introduction of cleaner technologies in Switzerland has let to reductions in the emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition. We infered different drivers of community change (i.e. Nitrogen deposition, climate warming, land-use change) in Swiss mountain hay meadows using data from the Swiss biodiversity monitoring.

@@

# Introduction {-}
Nitrogen deposition is the entry of reactive nitrogen compounds into water and soil via deposition through the air. Since Nitrogen is an essential plant nutrient and many species-rich communities are adapted to conditions of low Nitrogen availability [@Vitousek1997], the addition of Nitrogen is likely to alter these communities. Indeed, Nitrogen deposition toghether with land-use and climate change is one of the major threats to biodiversity [@Sala2000; @Bobbink2010]. While there is strong evidence for the reduction in diversity of species-rich grasslands due to increased Nitrogen deposition [@Stevens2006; @Dupre2010; @Maskell2010; @Wesche2012], mountain grasslands have received less attention [@Humbert2016].

Since the late 1980s measures to reduce atmospheric pollution have successfully reduced emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition in many parts of Europe [@Torseth2012; @Fowler2007]. However, there are a number of factors that may prevent the recovery of plant communities that sufered from increased Nitrogen deposition. First, even if Nitrogen deposition is reduced, large areas might still be above the critical load [@Slootweg2015; @RihmAchermann2016]. Second, there is a potential that communities reach an alternative stable state after decades of increased Nitrogen deposition and these species are unlikely to disappear again after the reduction of Nitrogen depositon [@Stevens2016]. Therefore, it is an open question whether and how fast the reduction in Nitrogen deposition rates will lead to the recovery of extant plant communities.

Recovry of extant plant communities from high Nitrogen depositon would imply that the state of communities measured at different time points is improving over time (i.e. improving biodiversity endpoints sensu @Rowe2017). Species richness is an easy to measure and communicate biodiversity entpoint and it is often negatively correlated with Nitrogen deposition [@Maskell2010; @Field2014; @Rowe2017]. Other metrics that are potentially more useful to reflect favourable changes can be derived from the traits of the species in a community. In Europe, environmental preference has often been expressed using incicator values assigned to each plant species [@Ellenberg1992]. Examples of such metrics would be the number of oligotrophic species or the average indicator value of the species in a community [@Roth2013; @Rowe2017]. However, the lack of a temporal trend in such biodiversity endpoints -- particularly in species richness -- does not necessarily mean that species composition remains unchanged. This is because immigration and extinction might be equally frequent and may cancel each other out [@Hillebrand2018]. Thus an useful approach to understanding biodiversity change is through estimates of biodiversity turnover reflecting both colonization and local extinction [@Hillebrand2018], especially if colonization and local extinction is compared to random expectation [@Chase2011].

In Switzerland, grassland accounts for 70 per cent of the agricultural land. With extensive cultivation, permanent grassland has a very high biodiversity. This applies in particular to the meadows in the alpine region, where meadows with a high plant diversity are of great importance agronomically (Leiber et al. 2005), for the image of the products and for the conservation of biodiversity (Hohl, 2006). Observation data of vascular plants from field recordings shows negative correlations of Nitrogen deposition with species richness and species composition in mountain hay meadows [@Roth2013]. Between 1990 and 2010 NOx emissions in Switzerland decreased by 46% and NH3 emissions by 14% [@MaasGrennfelt2016], suggesting that Nitrogen deposition has decreased also in mountain hay meadows and we expected plant communities to respond accordingly.

However, Swiss mountain ecosystems are threatened also by other global change drivers. In Switzerland, temperatures increased from 1959 to 2008 at all altitudes, with an average warming rate of 0.35$^{\circ}$C per decade, which is about 1.6 times the northern hemispheric warming rate [@Ceppi2012]. This climate warming is likely to directly affect or interact with Nitrogen deposition to drive plant community changes [@Humbert2016]. Indeed, in an earlier study we found that at a landscape scale plant communities responded to climate warming even within a comparably short time period [@Roth2014] and particularly the shift of plant communities at mountain summits are the restult of recent climate warming interacting with airborne nitrogen deposition [@Rumpf2018](>> Steinbauer<<). Furthermore, traditional management regimes are currently changing wich also has major impacts on plant communities in mountainous regions of Europe [@Niedrist2009; @Homburger2012]. Management regimes of accessible mountainous areas are frequently intensified, while poorly accessible mountainous areas are being abandoned [@Tasser2002; @Strebel2015].

Here, we used the data from the Swiss biodiversity monitoring [@Weber2004] and addressed the following questions: (1) Did biodiversity endpoints that are likely to reflect temperature, precipitation, Nitrogen deposition or land-use intensity change over the last 15 years, (2) Was species turnover correlated with the avearage temperature, precipitation, Nitrogen deposition or stepness (we expect step areas to be less intensively managed), and (3) did species that newly colonized or disappeard (local extinction) from sites different from random expectation according to their incicator values for tempearture, humidity, nutrients or light.

@@

## Monitoring data and community measures{-}
We analysed the presence/absence of vascular plants sampled within the scope of Switzerland's Biodiversity Monitoring (BDM) programme that was launched in 2001 to monitor Switzerland's biodiversity and to comply with the Convention on Biological Diversity of Rio de Janeiro [@Weber2004]. The sampling sites were circles with a size of 10 m$^2$ and data collection was carried out by qualified botanists who visited each sampling site twice within the same season. During each visit all the vascular plant species detected on the plot were recorded. After the sampling of the plant data the botanists also assigned a habitat type to each sampling site according to the classification system developed for Switzerland [@Delarze2008].

We matched the habitat types of the Swiss classification system with the categories from the EUNIS system [level-3 classification; @Davies2004] and selected all sampling sites in mountain hay meadows (EUNIS E2.3). For more details on the field methods see @Plattner2004, @Roth2013 and @Roth2017. We analysed the data from 2003 to 2017. During that study period each sampling site was surveyed once per five-year period: the first period lasted from 2003 to 2007, the second from 2008 to 2012 and the third from 2013 to 2017. These selection criteria resulted in sample of `r nrow(sites)` sites.

For each survey (a survey consists of the two visits per season) we calculated the following community measures: (1) The number of recorded species (species richness), the community mean of indicator values of recorded species for  temperature (2), humidity (3), nutrients (4) and light (5). The indicator values were obtained from the recalibrated indicator values for the Swiss Flora [@Landolt2010]. Additionally to these five community measures that describe the state of plant communities for each site at a given time point, we also estimated the temporal turnover (i.e. species exchange ratio sensu @Hillebrand2018) as the proportion of species that differ between two time points to describe the community change between two time points.

To test data quality, independent replicate surveys are routinely performed within the BDM programm by botansists who were not involved in the regular BDM surveys. The regular surveyors did not know if and which sites were replicated [@Plattner2004]. We used the data from `r nrow(turnover_double)` such replicated surveys that were available to calculate the pseudo-turnover, which is the proportion of species that differed between two surveys that were conducted by two different surveyors during the same year on the same site.

## Environmental gradients {-}
We expected different global change patterns to simultaneously cause temporal change in mountain hay communities. To better disentangle the importance of these mechanisms we ordered the sites along four main environmental gradients. First we expected communities to response to climate warming and the response might be different depending along the temperature gradient [@Roth2014]. To describe the temperature gradient we used the mean annual temperature per site from the WordClim database [@Fick2017]. Another key drivers for plants that is likely to be affected by climate change is precipitation [@Beier2012]. We used the annual precipitation per site from the WordClim database [@Fick2017]. Further, we estimated atmospheric Nitrogen deposition for each site using a pragmatic approach described in @Rihm2001 that combines monitoring data, spatial interpolation methods, emission inventories, statistical dispersion models and inferential deposition models. For details on the estimation of Nitrogen deposition see @RihmAchermann2016 and @Roth2017. Finally we used inclination as proxy for land-use intensity because we expected that steeper sites are likely to be less intensively managed [@Strebel2015].

## Statistical analyses {-}
To estimate the linear trend over time for each of the five community measures we applied linear mixed models (LMM) with normal distribution except for species-richness with Poisson distribution and the logarithm as link function. We specified site specific trends with the assumption that the between-site differences in intercepts and slopes can be described with normal distributions [i.e. a random intercept random slope model, @German2006]. Model parameters were estimated in a Bayesian framework using the R-Package *rstanarm* [@Stan2016, @Muth2018].

@@

To infer whether species that colonized or disappeared from a site had particular indicator values that differed from the other species at that site, we produced for each site a list with all species that were recorded during the three surveys (total community). We then calculated the community mean (CM) of the indicator value for all species that colonized the site during the three surveys (i.e. not recorded during first survey and recorded during second or not recorded during second and recorded during third). We then randomly selected the same number of species from the total community and also calculated the community mean of the value for these species (random-CM). We repeated the random selection of species 1000 times. Then we calculated the differences of the CM minus the average of the random-CMs to obtain a standardized measure (standardized-CM) of how different the disappearing species were from random expectation. A standardized-CM < 0 would suggest that the indicator value of colonizing species were lower than the average species that were occurring at this site. We applied this method for both colonizing and disappearing species and for the indicator values for temperature, humidity, nutrients and light (see Appendix A). We then inferred whether the standardized-CM is changing along the corresponding gradient using linear models. Model parameters were estimated in a Bayesian framework using the R-Package *rstanarm* [@Stan2016, @Muth2018].

We used logistic-LMM to infer whether the colonization probability or local survival probability was changing along the Nitrogen deposition gradient and whether this changing depended on the species indicator value for nutrients. To analyse the colonization probability we selected all species that were not observed during the first survey, and asked if they were observed (then Occ = 1) or not observed (then Occ = 0) during the second survey. The same was also done for all species that were not observed during the second survey. The variable 'occ' was then used as dependent variable in the logistic-LMM. As predictor variables the model contained the Nitrogen deposition of the site, the indicator value for nutrients of the species and the interaction of these two variables. Additionally  species-ID and site-ID were included as random effects. The same logistic_LMM was also used to infer local survival probability. In that case, however, we selected all species that were recorded during the first or second survey and the variable 'occ' then indicated whether or not the species was also observed during the next survey. Model parameters were estimated using an approximate Bayesian approach using the R-Package *arm* [@arm2018].

@@

```{r communitytrendstab, echo=FALSE}
options(knitr.kable.NA = '') 
kable(res, align = c("l", rep("r", 7)), digits = 2,  booktabs = T, col.names = c("Measures", "Period 1", "Period 2", "Period 3", "Trend", "5%", "95%", "Prob. for trend"), caption = "Average measures of community structure for the three sampling periods (period 1: 2003-2007; period 2: 2008-2012; period: 2013-2017). The temporal trends are given as change per 10 years and were estimated from linear mixed models with normal distribution (except for species richness with Poisson distribution and a log-link function). The measure of precision for the temporal trend is given as the 5\\% and 95\\% quantiles of the marginal posterior distribution of the linear trend. The column 'Prob. for trend' gives the probability that the linear trend is $> 0$.")
```

The five measures of plant community structure suggested that plant communities in mountain hay meadows were rather stable between 2003 and 2017 and did not show a clear increase or decrease over time (Table \@ref(tab:communitytrendstab)): for each of the three 5-year survey periods the averages of species-richness and the average indicator values for temperature, humidity, nutrients and light did not vary much among the three sampling periods and the estimated trends were rather small. Except for average indicator value for temperature, the 90% credible-interval of the temporal trend contained zero. The results from the linear mixed models suggest that a linear temporal change was most likely for the community mean of the indicator value for temperature (probability of increase: `r 
res %>% filter(Measures == "Temperature value") %>% pull("Prob. for Trend")`), followed by the community mean of the indicator value for light (probability of decrease: `r 1 - res %>% filter(Measures == "Light value") %>% pull("Prob. for Trend")`) and it was least likely for the species richness (probability of increase: `r res %>% filter(Measures == "Species-richness") %>% pull("Prob. for Trend")`). The chance that the community mean of the nutrient value decreased between 2003 and 2017 was `r 1- res %>% filter(Measures == "Nutrients value") %>% pull("Prob. for Trend")`.

@@

```{r difffromrandomtab}
kable(res, align = c("l", "l", rep("c", 6)), digits = 3,  booktabs = T, col.names = c("Ellenberg value", "Gradient", "Estimate", "5%", "90%", "Estimate", "5%", "90%"), caption = "Difference in the average indicator value of species that (a) disappeared from site or (b) newly colonized a site compared to the same number of species that were randomly selected from all species recorded at a site. Shown are the results from linear model with the difference between disappeard/colonized species and random species as dependend variable and the sitemeasure (gradient) as predictor variable.") %>% 
  kable_styling() %>% 
  group_rows("(a) Plants that disappeard from a site", 1, 4, bold = F, italic = T) %>%
  group_rows("(b) Plants that newly colonized a site", 5, 8, bold = F, italic = T) %>%
  add_header_above(c(" " = 2, "Difference from random" = 3, "Change along gradient" = 3))
```

High species turnover at a site is the result of species that disappeared from the site and species that newly colonized the site. To better understand the factors that drive these changes we are particularly interested whether the species that disappeared or colonized the sites differed in indicator values compared to what would be expected if the same number of species randomly disappeared or colonized the sites (i.e. random disappearance and random colonization) and whether there is a change along the gradients. It seems that the indicator values of newly colonizing species differed more from random colonization than the indicator values of disappearing species (Table \@ref(tab:difffromrandomtab)). For colonizing species, we found the largest differences from random colonization in the indicator value for nutrients: at sites with nitrogen deposition of 10 kg ha$^{-1}$ yr$^{-1}$ the newly colonizing species had in average a lower indicator value for nutrients than species under random disappearance (column "Difference from random" in Table \@ref(tab:difffromrandomtab)), but this differences between colonizing species and random colonization decreased with increasing Nitrogen deposition (column "Change along gradient" in Table \@ref(tab:difffromrandomtab)). Thus at high Nitrogen deposition colonizing species did not differ from random species (see Figure 3b in Appendix A).

While colonizing species had higher temperature values compared to what we would expect under random colonization, the differences between colonizing species and random colonization was about four times smaller compared to the difference in indicator value for nutrients between colonizing species and random species. Nevertheless, the variation in indicator value for temperature seemed important to explain the total species turnover. This is because, disappearing species tend to have lower temperature value than random species as well as colonizing species tend to have higher temperature values than random species; both processes lead to an overall replacement of species with lower temperature value with species with higher temperature values. This was not the case for the indicator value for nutrients: species with lower nutrients values tended to be more likely to disappear from as well as to colonize sites compared to random species (Table \@ref(tab:difffromrandomtab)). See also Appendix A were we present detailed results for the comparison between colonizing or disappearing species with randomly selected species.

@@

```{r cl, fig.cap="Colonization (a) and local survival (b) of oligotrophic (indicator value for nutrients <3; red line) and eutrophic (indicator value for nutrients >3) species along the Nitrogen deposition gradient. Given are means and 95\\%-Credible Intervals from logistic linear mixed models. The vertical lines indicat the deposition rate with equal colonization or survival probabilities for oligotrophic and eutrophic species with the solid line indicating the median and the dashed lines the 5\\% and 95\\% quantiels of the margional posterior distribution."}
X <- 1:30
Xsd <- (X - 10) / 10

@@

In Fig. \@ref(fig:cl) we compare the colonization and local survival probability of oligotrophic (indicator value of nutrients <3) and eutrophic (indicator value of nutrients > 3) species along the Nitrogen deposition gradient. Local survival probability was the same for oligotrophic and euthrophic species at a deposition rate of `r CL_surv` kg N ha$^{-1}$ yr$^{-1}$; colonization probability was the same for for oligotrophic and euthrophic species at a deposition rate of `r CL_col` kg N ha$^{-1}$ yr$^{-1}$. In only `r mean((sites$NTOT * 10 + 10) < 12.5)`% of the sites the deposition rate was below 12.5 kg N ha$^{-1}$ yr$^{-1}$ where the replacement of eutrophic with oligotrophic species is likely.

While we could not detect a consistent decrease in average indicator value for nutrients (Table \@ref(tab:communitytrendstab)), the higher colonization rate of species with low nutrient value at sites with low deposition rate seems to affect the spatial variation of species richness: sites with low Nitrogen deposition are likely to become more species rich over time likely resulting in steeper slope of the negative relationship between Nitrogen deposition and species richness. Indeed, if we apply at different time points a similar model as in @Roth2013 to infer the effects of Nitrogen deposition on the spatial variation of species richness, the resulting effect size (i.e. the slope) becomes more negative over time (Fig. \@ref(fig:figconsequences)).

```{r figconsequences, out.width = '50%', fig.asp=0.8, fig.width=4, fig.cap="Effect size of Nitrogen deposition on total species richness estimated from applying the Poisson-GLM with species richness as dependend variable and Nitrogen deposition plus other site covariates as predictors using only the surveys from one five-year interval. Note that within every five-year interval all plots were sampled once."}
# Add site covariates to 'surv'
dat <- surv %>% left_join(sites)

>>>>

affiliation:
  - code: 1
    address: Zoological Institute, University of Basel, Basel, Switzerland
  - code: 2
    address: Hintermann  Weber AG, Austrasse 2a, 4153 Reinach, Switzerland
  - code: 3
    address: Meteotest, Fabrikstrasse 14, 3012 Bern, Switzerland
  - code: 4
    address: Agroscope, Swiss Soil Monitoring Network NABO, 8046 Zurich, Switzerland
  - code: 5
    address: Federal Office for the Environment (FOEN), Air Pollution Control and Chemicals Division, 3003 Bern, Switzerland
abstract: >
  Nitrogen (N) deposition is a major threat to biodiversity of many habitats. The recent introduction of cleaner technologies in Switzerland has let to reductions in the emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition. We inferred different drivers of community change (i.e. Nitrogen deposition, climate warming, land-use change) in Swiss mountain hay meadows using data from the Swiss biodiversity monitoring. While oligotrophic species were more likely to colonize compared to random expectation, we found only weak shifts in plant community structure. Particularly, the average nutrient value of plant communities remained stable over time. Apparently, the community effects of the replacement of eutrophic species with oligotrophic species was outweighed by climate warming. Our results add to the increasing evidence that inferring species turnover will generate a far more reliable understanding of the biotic response to changing environments than solely tracking average community composition.

@@

# Introduction {-}
Nitrogen deposition is the entry of reactive nitrogen compounds into water and soil via deposition through the air. Since Nitrogen is an essential plant nutrient and many species-rich communities are adapted to conditions of low Nitrogen availability [@Vitousek1997], the addition of Nitrogen is likely to alter these communities. Indeed, Nitrogen deposition together with land-use and climate change is one of the major threats to biodiversity [@Sala2000; @Bobbink2010]. While there is strong evidence for the reduction in diversity of species-rich grasslands due to increased Nitrogen deposition [@Stevens2006; @Dupre2010; @Maskell2010; @Wesche2012], mountain grasslands have received less attention [@Humbert2016].

Since the late 1980s measures to reduce atmospheric pollution have successfully reduced emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition in many parts of Europe [@Torseth2012; @Fowler2007]. However, there are a number of factors that may prevent the recovery of plant communities that suffered from increased Nitrogen deposition. First, even if Nitrogen deposition is reduced, large areas might still be above the critical load [@Slootweeg2015; @RihmAchermann2016]. Second, there is a potential that communities reach an alternative stable state after decades of increased Nitrogen deposition and these species are unlikely to disappear again after the reduction of Nitrogen deposition [@Stevens2016]. Therefore, it is an open question whether and how fast the reduction in Nitrogen deposition rates will lead to the recovery of extant plant communities.

Recovery of extant plant communities from high Nitrogen deposition would imply that the state of communities measured at different time points is improving over time (i.e. improving biodiversity endpoints sensu @Rowe2017). Species richness is an easy to measure and communicate biodiversity endpoint and it is often negatively correlated with Nitrogen deposition [@Maskell2010; @Field2014; @Rowe2017]. Other metrics that are potentially more useful to reflect favourable changes can be derived from the traits of the species in a community. In Europe, environmental preference has often been expressed using indicator values assigned to each plant species [@Ellenberg1992]. Examples of such metrics would be the number of oligotrophic species or the average indicator value of the species in a community [@Roth2013; @Rowe2017]. However, the lack of a temporal trend in such biodiversity endpoints -- particularly in species richness -- does not necessarily mean that species composition remains unchanged. This is because immigration and extinction might be equally frequent and may cancel each other out [@Hillebrand2018]. Thus a useful approach to understanding biodiversity change is through estimates of biodiversity turnover reflecting both colonization and local extinction [@Hillebrand2018], especially if colonization and local extinction is compared to random expectation [@Chase2011].

In Switzerland, grassland accounts for 70% of the agricultural land. With extensive cultivation, permanent grassland has a very high biodiversity. This applies in particular to the meadows in the alpine region, where meadows with a high plant diversity are also of importance agronomically [@Leiber2006]. In mountain hay meadows, the spatial variation of species richness in vascular plants is negative correlated with Nitrogen deposition [@Roth2013], suggesting that mountain grasslands suffered from increased Nitrogen deposition. However, between 1990 and 2010 NO$_x$ emissions in Switzerland decreased by 46% and NH$_3$ emissions by 14% [@MaasGrennfelt2016] and we expect plant communities to recover.

However, Swiss mountain ecosystems are threatened also by other global change drivers. In Switzerland, temperatures increased from 1959 to 2008 at all altitudes, with an average warming rate of 0.35$^{\circ}$C per decade, which is about 1.6 times the northern hemispheric warming rate [@Ceppi2012]. This climate warming is likely to directly affect or interact with Nitrogen deposition to drive plant community changes [@Humbert2016]. Indeed, in an earlier study we found that at the landscape scale plant communities responded to climate warming even within a comparably short time period [@Roth2014] and particularly the shift of plant communities at mountain summits are the result of recent climate warming interacting with airborne nitrogen deposition [@Steinbauer2018]. Furthermore, traditional management regimes are currently changing, which also has major impacts on plant communities in mountainous regions of Europe [@Niedrist2009; @Homburger2012]. Management regimes of accessible mountainous areas are frequently intensified, while poorly accessible mountainous areas are being abandoned [@Tasser2002; @Strebel2015].

Here, we used the data from the Swiss biodiversity monitoring [@Weber2004] and addressed the following questions: (1) Did biodiversity endpoints that are likely to reflect temperature, precipitation, Nitrogen deposition or land-use intensity change over the last 15 years, (2) was species turnover correlated with the average temperature, precipitation, Nitrogen deposition or inclination (we expect steep areas to be less intensively managed), and (3) did species that newly colonized or disappeared (local extinction) from sites differ from random expectation according to their indicator values for temperature, humidity, nutrients or light.

@@

## Monitoring data and community measures {-}
We analysed the presence/absence of vascular plants sampled within the scope of Switzerland's Biodiversity Monitoring (BDM) programme that was launched in 2001 to monitor Switzerland's biodiversity and to comply with the Convention on Biological Diversity of Rio de Janeiro [@Weber2004]. The sampling sites were circles with a size of 10 m$^2$ and data collection was carried out by qualified botanists who visited each sampling site twice within the same season. During each visit all the vascular plant species detected on the plot were recorded. For details on the field methods see @Plattner2004, @Roth2013 and @Roth2017.

After the sampling of the plant data the botanists also assigned a habitat type to each sampling site according to the classification system developed for Switzerland [@Delarze2008]. We matched the habitat types of the Swiss classification system with the categories from the EUNIS system [level-3 classification; @Davies2004] and selected all sampling sites in mountain hay meadows (EUNIS E2.3). We analysed the data from 2003 to 2017. During that study period each sampling site was surveyed once per five-year period: the first period lasted from 2003 to 2007, the second from 2008 to 2012 and the third from 2013 to 2017. These selection criteria resulted in sample of `r nrow(sites)` sites.

For each survey (a survey consists of the two visits per season) we calculated the following biodiversity endpoints [@Rowe2017]: (1) The number of recorded species (species richness), the community mean of indicator values of recorded species for (2) temperature, (3) humidity, (4) nutrients and (5) light. The indicator values were obtained from the recalibrated indicator values for the Swiss Flora [@Landolt2010]. Additionally, to these five biodiversity endpoints that describe the state of plant communities for each site at a given time point, we also estimated the temporal turnover (i.e. species exchange ratio sensu @Hillebrand2018) as the proportion of species that differ between two time points to describe the community change between two time points.

To test data quality, independent replicate surveys are routinely performed within the BDM program by botanists who were not involved in the regular BDM surveys. The regular surveyors did not know if and which sites were replicated [@Plattner2004]. We used the data from `r nrow(turnover_double)` such replicated surveys that were available to calculate the pseudo-turnover, which is the proportion of species that differed between two surveys that were conducted by two different surveyors during the same year on the same site.

## Environmental gradients {-}
We expected different global change patterns to simultaneously cause temporal change in mountain hay communities. To better disentangle the importance of these mechanisms we ordered the sites along four main environmental gradients. First, we expected communities to response to climate warming and the response might be different along the temperature gradient [@Roth2014]. To describe the temperature gradient, we used the mean annual temperature per site from the WordClim database [@Fick2017]. Another key drivers for plants that is likely to be affected by climate change is precipitation [@Beier2012]. We used the annual precipitation per site from the WordClim database [@Fick2017]. Further, we estimated atmospheric Nitrogen deposition for each site using a pragmatic approach described in @Rihm2001 that combines monitoring data, spatial interpolation methods, emission inventories, statistical dispersion models and inferential deposition models. For details on the estimation of Nitrogen deposition see also @RihmAchermann2016 and @Roth2017. Finally, we used inclination as proxy for land-use intensity because we expected that steeper sites are likely to be less intensively managed [@Strebel2015].

## Statistical analyses {-}
To estimate the linear trend over time for each of the five biodiversity endpoints we applied linear mixed models (LMM) with normal distribution except for species-richness with Poisson distribution. We specified site specific trends with the assumption that the between-site differences in intercepts and slopes can be described with normal distributions [i.e. a random intercept random slope model, @Gelman2006]. Model parameters were estimated in a Bayesian framework using the R-Package *rstanarm* [@Stan2016, @Muth2018].

@@

To infer whether species that colonized or disappeared from a site had particular indicator values that differed from the other species at that site, we produced for each site a list with all species that were recorded during the three surveys (total community). We then calculated the community mean (CM) of the indicator value for all species that colonized the site during the three surveys (i.e. not recorded during first survey and recorded during second or not recorded during second and recorded during third). We then randomly selected the same number of species from the total community and also calculated the community mean of the value for these species (random-CM). We repeated the random selection of species 1000 times. Then we calculated the differences of the CM minus the average of the random-CMs to obtain a standardized measure (standardized-CM) of how different the disappearing species were from random expectation. A standardized-CM < 0 would suggest that the indicator values of colonizing species were lower than the indicator values of average species that were occurring at this site. We applied this method for both colonizing and disappearing species and for the indicator values for temperature, humidity, nutrients and light (see Appendix A). We then inferred whether the standardized-CM is changing along the corresponding gradient using linear models. Model parameters were estimated in a Bayesian framework using the R-Package *rstanarm* [@Stan2016, @Muth2018].

We used logistic-LMM to infer whether the colonization probability or local survival probability was changing along the Nitrogen deposition gradient and whether this changing depended on the species indicator value for nutrients. To analyse the colonization probability, we selected all species that were not observed during the first survey and asked if they were observed (then Occ = 1) or not observed (then Occ = 0) during the second survey. The same was also done for all species that were not observed during the second survey if they were observed or not observed (then Occ = 0) during the third survey. The variable 'occ' was then used as dependent variable in the logistic-LMM. As predictor variables the model contained the Nitrogen deposition of the site, the indicator value for nutrients of the species and the interaction of these two variables. Additionally, species-ID and site-ID were included as random effects. The same logistic LMM was also used to infer local survival probability. In that case, however, we selected all species that were recorded during the first or second survey and the variable 'occ' then indicated whether or not the species was also observed during the next survey. Model parameters were estimated using an approximate Bayesian approach using the R-Package *arm* [@arm2018].

@@

```{r communitytrendstab, echo=FALSE}
options(knitr.kable.NA = '') 
kable(res, align = c("l", rep("r", 7)), digits = 2,  booktabs = T, col.names = c("Measures", "Period 1", "Period 2", "Period 3", "Trend", "5%", "95%", "Prob. for trend"), caption = "Average measures of the biodiversity endpoints for the three sampling periods (period 1: 2003-2007; period 2: 2008-2012; period: 2013-2017). The temporal trends are given as change per 10 years and were estimated from linear mixed models with normal distribution (except for species richness with Poisson distribution and a log-link function). The measure of precision for the temporal trend is given as the 5\\% and 95\\% quantiles of the marginal posterior distribution of the linear trend. The column 'Prob. for trend' gives the probability that the linear trend is $> 0$.")
```

The five measures of plant community structure (i.e. biodiversity endpoints) suggested that plant communities in mountain hay meadows were rather stable between 2003 and 2017 and did not show a clear increase or decrease over time (Table \@ref(tab:communitytrendstab)): for each of the three 5-year survey periods the averages of species-richness and the average indicator values for temperature, humidity, nutrients and light did not vary much among the three sampling periods and the estimated trends were rather small. Except for average indicator value for temperature, the 90% credible-interval of the temporal trend contained zero. The results from the linear mixed models suggest that a linear temporal change was most likely for the community mean of the indicator value for temperature (probability of increase: `r 
res %>% filter(Measures == "Temperature value") %>% pull("Prob. for Trend")`), followed by the community mean of the indicator value for light (probability of decrease: `r 1 - res %>% filter(Measures == "Light value") %>% pull("Prob. for Trend")`) and it was least likely for the species richness (probability of increase: `r res %>% filter(Measures == "Species-richness") %>% pull("Prob. for Trend")`). The chance that the community mean of the nutrient value decreased between 2003 and 2017 was `r 1- res %>% filter(Measures == "Nutrients value") %>% pull("Prob. for Trend")`.

@@

```{r difffromrandomtab}
kable(res, align = c("l", "l", rep("c", 6)), digits = 3,  booktabs = T, col.names = c("Ellenberg value", "Gradient", "Estimate", "5%", "90%", "Estimate", "5%", "90%"), caption = "Difference in the average indicator value of species that (a) disappeared from site or (b) newly colonized a site compared to the same number of species that were randomly selected from all species recorded at a site. Shown are the results from linear model with the difference between disappeared/colonized species and random species as dependent variable and the site-measure (gradient) as predictor variable.") %>% 
  kable_styling() %>% 
  group_rows("(a) Plants that disappeard from a site", 1, 4, bold = F, italic = T) %>%
  group_rows("(b) Plants that newly colonized a site", 5, 8, bold = F, italic = T) %>%
  add_header_above(c(" " = 2, "Difference from random" = 3, "Change along gradient" = 3))
```

High species turnover at a site is the result of species that disappeared from the site and species that newly colonized the site. To better understand the factors that drive these changes we were particularly interested whether the species that disappeared or colonized the sites differed in indicator values compared to what would be expected if the same number of species randomly disappeared or colonized the sites (i.e. random disappearance and random colonization) and whether there is a spatial variation along the gradients. It seems that the indicator values of newly colonizing species differed more from random colonization than the indicator values of disappearing species (Table \@ref(tab:difffromrandomtab)). For colonizing species, we found the largest differences from random colonization in the indicator value for nutrients: at sites with nitrogen deposition of 10 kg ha$^{-1}$ yr$^{-1}$ the newly colonizing species had in average a lower indicator value for nutrients than species under random disappearance (column "Difference from random" in Table \@ref(tab:difffromrandomtab)), but this differences between colonizing species and random colonization decreased with increasing Nitrogen deposition (column "Change along gradient" in Table \@ref(tab:difffromrandomtab)). Thus, at high Nitrogen deposition colonizing species did not differ from random species (see Figure 3b in Appendix A).

While colonizing species had higher temperature values compared to what we would expect under random colonization, the differences between colonizing species and random colonization was about four times smaller compared to the difference in indicator value for nutrients between colonizing species and random species. Nevertheless, the variation in indicator value for temperature seemed important to explain the total species turnover. This is because, disappearing species tend to have lower temperature value than random species as well as colonizing species tend to have higher temperature values than random species; both processes lead to an overall replacement of species with lower temperature value with species with higher temperature values. This was not the case for the indicator value for nutrients: species with lower nutrients values tended to be more likely to disappear from as well as to colonize sites compared to random species (Table \@ref(tab:difffromrandomtab)). See also Appendix A where we present detailed results for the comparison between colonizing or disappearing species with randomly selected species.

@@

```{r cl, fig.cap="Colonization (a) and local survival (b) of oligotrophic (indicator value for nutrients <3; red line) and eutrophic (indicator value for nutrients >3) species along the Nitrogen deposition gradient. Given are means and 95\\%-Credible Intervals from logistic linear mixed models. The vertical lines indicate the deposition rate with equal colonization or survival probabilities for oligotrophic and eutrophic species with the solid line indicating the median and the dashed lines the 5\\% and 95\\% quantiles of the marginal posterior distribution."}
X <- 1:30
Xsd <- (X - 10) / 10

@@

In Fig. \@ref(fig:cl) we compare the colonization and local survival probability of oligotrophic (indicator value of nutrients <3) and eutrophic (indicator value of nutrients > 3) species along the Nitrogen deposition gradient. Local survival probability was the same for oligotrophic and euthrophic species at a deposition rate of `r CL_surv` kg N ha$^{-1}$ yr$^{-1}$; colonization probability was the same for oligotrophic and eutrophic species at a deposition rate of `r CL_col` kg N ha$^{-1}$ yr$^{-1}$. In only `r mean((sites$NTOT * 10 + 10) < 12.5)`% of the sites the deposition rate was below 12.5 kg N ha$^{-1}$ yr$^{-1}$ where the replacement of eutrophic with oligotrophic species is likely.

While we could not detect a consistent decrease in average indicator value for nutrients (Table \@ref(tab:communitytrendstab)), the higher colonization rate of species with low nutrient value at sites with low deposition rate seems to affect the spatial variation of oligotrophic species richness: sites with low Nitrogen deposition are likely to become more species rich over time likely resulting in steeper slope of the negative relationship between Nitrogen deposition and oligotrophic species richness. Indeed, if we apply at different time points a similar model as in @Roth2013 to infer the effects of Nitrogen deposition on the spatial variation of oligotrophic species richness, the resulting effect size (i.e. the slope) became more negative over time (Fig. \@ref(fig:figconsequences)).

```{r figconsequences, out.width = '50%', fig.asp=0.8, fig.width=4, fig.cap="Effect size of Nitrogen deposition on oligotrophic species richness estimated from applying the Poisson-GLM with species richness as dependent variable and Nitrogen deposition plus other site covariates as predictors using only the surveys from one five-year interval. Note that within every five-year interval all plots were sampled once."}
# Add site covariates to 'surv'
dat <- surv %>% left_join(sites)
